### 2021-10-25 documents/sections/extensions.tex
new autogenerated tables

All available descriptions of the third Pekodian language, \ikpeng, list \obj{k-} as the only \gl{1}\gl{s_a_} marker \parencites[55]{ikpengpacheco1997}[105]{campetela1997analise}[64]{ikpengpacheco2001}[205]{alves2013verbo}.
However, most \ikpeng cognates of the \arara verbs with \gl{1}\gl{s_a_} \obj{w-} actually do not take \obj{k-}, but rather \obj{ɨ-} or Ø, as shown in \exref{ikpw}.
The exception is \qu{to go}, which has \obj{k-} \exref{ikp-170}.
There is a formally identical \ikpeng cognate of \arara \obj{iptoŋ} \qu{to go down}, but no first person forms are attested \perscommpar{Angela Chagas}.
While there are reflexes of \rc{a[p]} \qu{to be} in \ikpeng, only \rc{eti} forms are attested inflected for first person \parencite[401]{gildea2018reconstructing}.

@@

%\ex<ikp-182>\ikpeng \parencite[][320]{chagas2019ikpeng}\\
%\begingl
%\gla ∅-iptoŋ-lan//
%\glb \gl{3}\gl{s_a_}-sit.down-\gl{pst}//
%\glft \qu{She sat down.}//
%\endgl
%\xe
%
While clearly not a regular sound change, \PXin \rc{w} is attested as sometimes being lost in \ikpeng, as evidenced by the correspondences in \exref{ikpwloss}.
Thus, we suggest that the \ikpeng \gl{1}\gl{s_a_} marker \obj{ɨ-}/Ø is cognate with \arara \gl{1}\gl{s_a_} \obj{w(ɨ)-}, and was subject to phonological erosion of \rc{w}, a phenomenon which will also be seen in other languages discussed in following sections.

\ex<ikpwloss> %\arara-\ikpeng \obj{w}:Ø correspondences\\
\begin{tabular}[t]{@{}lll@{}}
\PXin & \arara & \ikpeng\\
\rc{wɨna} \qu{\gl{dat}} & \obj{wɨna} & \obj{ɨna}\\
\rc{wokori} \qu{dog, jaguar} & \obj{wokori} & \obj{akari}\\
\rc{tawe} \qu{capuchin monkey} & \obj{tawe} & \obj{tae}\\
\rc{watke} \qu{to defecate} & \obj{watke} & \obj{atke}\\
\rc{wɨnkɨ} \qu{to sleep} & \obj{wɨnkɨ} & \obj{ɨnkɨ}\\
& (\obj{o-wɨnkɨ} \qu{sleep!}) & (\obj{j-ɨnkɨ-lɨ} \qu{s/he slept})\\
\end{tabular}\\
\parencites[80]{alves2017arara}[164, 67, 150]{ikpengpacheco2001}[9]{desouza2010arara}[68]{souza1993arara}[91]{alves2013verbo}[40]{campetela1997analise} 
\xe

\begin{table}
	\centering
	\caption{Verbs preserving \gl{1}\gl{s_a_} in \PPek}
	\label{tab:pekverbs}
	\begin{tabular}{@{}lllll@{}}
	\mytoprule
& \PPek & \arara & \ikpeng & \bakairi \\
\mytoprule
\qu{say} & \rc{wɨ-ge-} & \obj{wɨ-ge-} & \obj{ɨ-ge-} & \obj{u-ge-} \\
\qu{bathe} & \rc{w-ipɨ-} & \obj{w-ibɨ-} & \obj{Ø-ip-} & \obj{w-i-} \\
\qu{be-1} & \rc{w-ap-} & \obj{w-ap-} & -- & \obj{w-a-} \\
\qu{be-2} & \rc{w-et͡ʃi-} & \obj{w-it͡ʃi-} & \obj{Ø-et͡ʃi-} & \obj{w-i-} \\
\qu{come} & \rc{w-epɨ}, \rc{k-əd-epɨ-} & \obj{w-ebɨ-} & \obj{k-arep-} & \obj{k-əewɨ-} \\
\qu{go down} & \rc{w-[ɨ/i]ptə-} & \obj{w-iptoŋ-} & ?\obj{-iptoŋ-} & \obj{k-ɨtəgɨ-} \\
\qu{go} & \rc{w-ɨtən-} & \obj{w-ɨdo-} & \obj{k-aran-} & \obj{u-tə-} \\
	\mybottomrule
	\end{tabular}
\end{table}

\Cref{tab:pekverbs} gives an overview of the first person forms of the seven verbs under discussion, along with our \PPek reconstruction.
The form for \rc{ge} \qu{to say} is straightforward to reconstruct; the \bakairi change of \rc{wɨ-} to \obj{u-} is not unexpected given correspondences like \bakairi \obj{udo} \parencite{meira2005southern} from \PC \rc{wɨtoto} \qu{person} \parencite[4]{gildea2007greenberg}.
\qu{to bathe} is straightforwardly reconstructible as \rc{ipɨ}.
For \qu{to be}, \ikpeng \obj{e} is very likely the original vowel, given the \PC form \rc{eti} \pcref{sec:be}.

@@

\pex<hixɨ>
\a<wai-108> \waiwai \parencite[][71]{waiwaihawkins1998}
\begingl
\glpreamble wɨɨkekɲe//
\gla wɨ-ka-jakɲe//
\glb \gl{1}-say-\gl{pst}//
\glft \qu{I said.}//
\endgl
\a<hix-118> \hixka \parencite[][124]{hixkaryanaderby1985}
\begingl
\glpreamble roxehra nay hamɨ Kaywerye ɨkekonɨ//
\gla ro-ʃe-hɨra n-a-je hamɨ kajwerʲe ɨ-ka-jakonɨ//
\glb \gl{1}-\gl{des}-\gl{neg} \gl{3}-be-\gl{npst}.\gl{uncert} \gl{evid} K. \gl{1}\gl{s_a_}-say-\gl{rem}.\gl{cont}//
\glft \qu{I said (to myself), “Kaywerye evidently doesn't like me”.}//
\endgl
\a<hix-17> \hixka \parencite[][191]{hixkaryanaderby1985}
\begingl
\gla ɨ-koroka-no//
\glb \gl{1}>\gl{3}-wash-\gl{imm}//
\glft \qu{I washed him.}//
\endgl
\a<wai-159> \waiwai \parencite[][192]{waiwaihawkins1998}
\begingl
\glpreamble wîyesî//
\gla wɨ-jo-jasɨ//
\glb \gl{1}>\gl{3}-boil-\gl{npst}//
\glft \qu{I will boil it.}//
\endgl
\xe
%
Thus, similar to the the case of \ikpeng, \hixka has lost the \rc{w} of the \envr{}{C} allomorph \rc{wɨ-}.
Notably, \textcite{hixkaryanaderby1985} does not see this \obj{ɨ-} as an irregular \gl{1}\gl{s_a_}, but as the regular \gl{1}>\gl{3} marker, because he considers \hixka \obj{ka} \qu{to say} to be transitive, an issue which will be discussed in \cref{sec:say}.

There are three more verbs which did not take innovative \rc{k-} in \PWai.
Two roots are part of the copula \rc{eʃi}/\obj{a(h)}, straightforwardly reconstructed in \exref{parpar.parcop}.
The other is \qu{to go} \exref{parpar.pargo}, somewhat of a special case.

\pex<parpar>
\a<parcop> Parukotoan \qu{to be}\\
\begin{tabular}[t]{@{}llll@{}}
& \PPar & \hixka & \waiwai \\
\gl{1}\gl{s_a_} & \rc{\emp{w-}eʃi-}/\obj{\emp{w-}ah-} & \obj{\emp{w-}eʃe-}/\obj{\emp{w-}ah-} & \obj{\emp{w-}eeʃi-}/\obj{\emp{w-}a-}\\
\gl{2}\gl{s_a_} & \rc{m-eʃi-}/\obj{m-ah-} & \obj{m-eʃe-}/\obj{m-ah-} & \obj{m-eeʃi-}/\obj{m-a-}\\
\gl{1+2}\gl{s_a_} & \rc{t-eʃi-}/\obj{t-ah-} & \obj{t-eʃe-}/\obj{t-ah-} & \obj{t͡ʃ-eeʃi-}/\obj{t-a-}\\
\gl{3}\gl{s_a_} & \rc{n-eʃi-}/\obj{n-ah-} & \obj{n-eʃe-}/\obj{n-ah-} & \obj{n-eeʃi-}/\obj{n-a-}\\
\end{tabular}\\
(\cite[197--198]{hixkaryanaderby1985} and Spike Gildea, p.c.)
\a<pargo> Parukotoan \qu{to go}\\
\begin{tabular}[t]{@{}llll@{}}
& \PPar & \hixka & \waiwai\\
\gl{1}\gl{s_a_} & \rc{\emp{wɨ-}to-} & \obj{\emp{ɨ-}to-} & \obj{kɨ\emp{w-}to-}\\
\gl{2}\gl{s_a_} & \rc{mɨ-to-} & \obj{mɨ-to-} & \obj{mɨɨ-to-}\\
\gl{1+2}\gl{s_a_} & \rc{tɨt-to-} & \obj{tɨ-to-} & \obj{tɨ-to-}/\obj{tɨh-t͡ʃe-}\\
\gl{3}\gl{s_a_} & \rc{nɨ-to-} & \obj{n-to-} & \obj{nɨɨ-to-}\\
\end{tabular}\\
(\cites[69--70, 211]{hixkaryanaderby1985}[179]{waiwaihawkins1998} and Spike Gildea, p.c.)
\xe
%
While \hixka has the expected \obj{ɨ-}, \waiwai seems to have combined innovative \gl{1}\gl{s_a_} \obj{k-} with the old \rc{w-}, an analysis also considered by \textcite[90]{gildea1998}.
Alternatively, this form was influenced by deverbalized forms of \qu{to go}, where a reflex of the \gl{s_a_} class marker \rc{w-} has become fossilized \exref{waiwto}.

@@

@@

As for verbs unaffected by the spread of \rc{t-}, \textcite{meira1998proto} lists the first four verbs in \exref{ptirw} as being reconstructible to \PTar with \rc{w-}, for which we offer our reconstructed \PTir forms.

@@

\pex<ptirw> \PTir \gl{1}\gl{s_a_} \rc{w-} \parencites[112--115]{meira1998proto}[294]{triomeira1999}\\ %[85]{gildea1994akuriyo}
\begin{tabular}[t]{@{}llll@{}}
& \PTir & \trio & \akuriyo\\
\qu{go} & \rc{wɨ-tə(mɨ)-} & \obj{wɨ-tə(mɨ)-} & \obj{ə-təmɨ-} \\% (\obj{wɨ-təemɨ-})\\
\qu{say} & \rc{wɨ-ka-} & \obj{wɨ-ka-} & \obj{wɨ-ka-} \\
\qu{come} & \rc{w-əʔepɨ-} & \obj{w-əepɨ-} & \obj{Ø-eepɨ-} \\
\qu{be} & \rc{w-ae-} & \obj{w-ae-} & Ø\obj{-aʔe-} \\
& \rc{w-eʔi-} & \obj{w-ei-} & ?-\obj{eʔi-}\\
\end{tabular}
%\a<aku12> \akuriyo \obj{kəʔ-eepɨ} \qu{we came} \parencite[114]{meira1998proto}
%\a<aku13> \akuriyo \obj{(w)i-toka} \qu{I hit him/her} \parencite[86]{gildea1994akuriyo}
%\a<tricome> \trio \obj{n-epɨ} \qu{s/he came} \parencite[114]{meira1998proto}
\xe

@@

\begin{table}
	\centering
	\caption{\akuriyo \gl{1}\gl{s_a_} markers in Gildea's fieldnotes}
	\label{tab:aku1sa}
	\begin{tabular}{@{}lll@{}}
	\mytoprule
first person \obj{k-} & first person \obj{t͡ʃ-} \\
\mymidrule
\obj{ənɨkɨ} \qu{to sleep} & \obj{eepɨ} \qu{to bathe} \\
\obj{əməmɨ} \qu{to enter} & \obj{ewai} \qu{to sit down} \\
\obj{əturu} \qu{to talk} & \obj{etonema} \qu{to lie down} \\
\obj{əət͡ʃena} \qu{to cry} & \obj{ekɨɨrɨka} \qu{to stay back} \\
\obj{ətajiŋka} \qu{to run} & \obj{entapo} \qu{to yawn} \\
\obj{əiwa} \qu{to tremble} &  \\
\obj{əempa} \qu{to learn} \\
	\mybottomrule
	\end{tabular}
\end{table}

%Let us first consider a scenario where \obj{k-} was the older \gl{1}\gl{s_a_} marker, occurring on both \obj{e}- and \obj{ə}-initial verbs.
%Next, \obj{t-}/\obj{t͡ʃ-} would have come in and replaced \obj{k-} before \obj{e}, but (mostly) not before \obj{ə}, resulting in the observable distribution.
%This scenario is not very satisfactory, since \obj{k-} is replaced in one completely unmotivated phonological environment, but not in the other.
%
%The other scenario is that \rc{t-} was the old first person marker occurring on the vast majority of \gl{s_a_} verbs.
%Subsequently, it developed into \rc{t͡ʃ-} / \_\obj{e}, but remained \rc{t-} / \_\obj{ə}; \obj{t͡ʃ}\goodtilde\obj{ʃ}\goodtilde\obj{s} and \obj{t} are phonemically contrastive in \akuriyo \parencite[16]{meira1998proto}.
%Thus, \obj{k-} can be seen as replacing \obj{t-}, but not \obj{t͡ʃ-}.
%The occurrences of \obj{t-} / \_\obj{ə} mentioned by \citeauthor{meira1998proto} are then remnants of the earlier prevalence of \obj{t-}.
%While this scenario also lacks a clear motivation, it at least explains the largely phonologically complementary distribution of \obj{k-} and \obj{t-} by the fact that the marker which was replaced had a phonemically different form than the one which was preserved.
%The other scenario which sees \rc{k-} as older would require \rc{t-} replacing \obj{k-} only in a specific environment.\footnote{Note that this process would be plausible even if the instances of \obj{t-} / \_\obj{ə} were due to \trio influence, as suggested by \textcite[108]{meira1998proto} -- these would then be a later reintroduction, replacing \obj{k-}, instead of remnants of the introduction of it.}

It has been speculated that the \trio \gl{s_a_} verb with a first person \obj{k-} marker, \obj{weka}/\obj{oeka} \qu{to defecate} might have something to do with \akuriyo \obj{k-}, which would potentially make the innovation of \rc{k-} a \PTir matter \parencite[116]{meira1998proto}.
However, this hypothesis faces a problem in that the first person form for \akuriyo \qu{to defecate} is \obj{j-ereina-} \parencite[88]{gildea1994akuriyo}.
Another hypothesis is that it originated in the corresponding \kaxui form \obj{ku-weka-} (form provided by Spike Gildea, p.c.), although the occurrence of \obj{o} in \trio would still need explanation.
%Rather, it seems likely that this irregular first person form originated in a class switch from \gl{s_p_} to \gl{s_a_} (\Cref{sec:resistant_verbs}), although it is unclear how exactly that switch resulted in the irregular form.

\subsection{\carijo \obj{j-}}
\label{sec:carijo}
As discussed by \textcite[105--107]{meira1998proto}, \carijo has extended the \gl{1}\gl{s_p_} marker \obj{j(i)-} to \gl{s_a_} verbs.\footnote{Since almost all \gl{s_a_} verbs are V-initial, only the \envr{}{V} allomorph \obj{j-} is attested.}
This, in combination with the extension of \gl{2}\gl{s_a_} \obj{m-} and \gl{1+2}\gl{s_a_} \obj{kɨt-}/\obj{kɨs-} to \gl{s_p_} verbs, results in a single unified \gl{s} category \exref{carpar}.

\ex<carpar> \carijo \parencites[106]{meira1998proto}[173]{robayo2000avance}\\
\begin{tabular}[t]{@{}lll@{}}
& \obj{tuda} \qu{to arrive} & \obj{eharaga} \qu{to dance}\\
\gl{1} & \obj{ji-tuda-} & \obj{j-eharaga-}\\
\gl{2} & \obj{mɨ-tuda-} & \obj{m-eharaga-}\\
\gl{1+2} & \obj{kɨsi-tuda-} & \obj{kɨs-eharaga-}\\
\gl{3} & \obj{ni-tuda-} & \obj{n-eharaga-}\\
\end{tabular}
\xe
%
Although the split-\gl{s} system has been lost entirely, former \gl{s_a_} verbs can be identified by taking a detransitivizer, like \obj{ehɨnəhɨ} \qu{to fight}, derived from \obj{hɨnəhɨ} \qu{to kill} \parencite[179]{robayo2000avance} \exref{car-30}.
%The same marker is also found in \gl{3}>\gl{1} scenarios \exref{car1p.car-1} and possessed nouns \exref{car1p.car-31}.

\ex<car-30>\carijo \parencite[][79]{koch1908hiana}\\
\begingl
\glpreamble hẹnẹ(x)tónoko-mā́ɽẹ y-e̱-hẹnẹ́(x)yai̯//
\gla hɨnəhtono-ko=marə j-e-hɨnəhɨ-jai//
\glb enemy-\gl{pl}=with \gl{1}-\gl{detrz}-kill-\gl{npst}.\gl{cert}//
\glft \qu{I fight with the enemies.}//
\endgl
\xe
%

%\pex<car1p>\carijo \parencite[][59, 53]{guerrero2016carijo}
%\begin{multicols}{2}
%\a<car-1>
%\begingl
%\gla jɨ-hɨnəh-ɨ//
%\glb \gl{3}>\gl{1}-kill-\gl{pfv}//
%\glft \qu{S/he killed me.}//
%\endgl
%\a<car-31>
%\begingl
%\gla j-owo-rɨ//
%\glb \gl{1}\gl{poss}-uncle-\gl{pert}//
%\glft \qu{my uncle}//
%\endgl
%\end{multicols}
%\xe
%

As noted in \cref{sec:taranoan}, this extension also erased any traces of a potential \PTar \gl{1}\gl{s_a_} marker \rc{t-}.
However, it did not fully eclipse the old \gl{1}\gl{s_a_} marker \rc{w-}, which is attested as being preserved in the verbs \obj{tə} \qu{to go} \exref{car1w.car-24} and \obj{a} \qu{to be} \exref{car1w.car-25}.

\pex<car1w>\carijo \parencite[][5, 42]{guerrero2016karihona}
\a<car-24>
\begingl
\gla wɨ-tə-e=rehe//
\glb \gl{1}-go-\gl{npst}=\gl{frust}//
\glft \qu{I almost go (but I am not going to go).}//
\endgl
\a<car-25>
\begingl
\gla əji-marə-ne w-a-e//
\glb \gl{2}-with-\gl{pl} \gl{1}-be-\gl{npst}//
\glft \qu{I am with you all.}//
\endgl
\xe
%
%Interestingly enough, the verb \qu{to be} only shows \obj{w-} in its \obj{a} root allomorph, but not in the \obj{et͡ʃi} root allomorph, where it has \obj{j-} \exref{car1y.mivida-12}.
%In a somewhat foreshadowing move, I will note that \PTar \rc{(ət)epɨ} \qu{to come} was also affected by the \carijo extension, as seen in \exref{car1y.car-18}.
%As discussed in \cref{sec:taranoan}, this verb preserves \rc{w-} in \akuriyo and \trio \exref{tri-105}.
%
%\pex<car1y>\carijo
%\a<mivida-12>
%\begingl
%\glpreamble iretibə et͡ʃinəme gərə jet͡ʃiɨ//
%\gla ireti-bə et͡ʃi-nə=me gərə j-et͡ʃi-ɨ//
%\glb then-from be-\gl{inf}=\gl{attrz} still \gl{1}-be-\gl{pfv}//
%\glft \qu{Then I was already grown up.} \parencite[][177]{robayo1989rame}//
%\endgl
%\a<car-18>
%\begingl
%\gla əji-wa-e j-eh-ɨ//
%\glb \gl{2}-search-\gl{sup} \gl{1}-come-\gl{pfv}//
%\glft \qu{I came looking for you.} \parencite[][102]{guerrero2019carijo}//
%\endgl
%\xe
%
%\ex<tri-105>\trio \parencite[][337]{triocarlin2004}\\
%\begingl
%\gla ene w-əe//
%\glb see.\gl{sup} \gl{1}\gl{s_a_}-come//
%\glft \qu{I came to see him/her.}//
%\endgl
%\xe

\subsection{\yukpa \obj{j-}}
\label{sec:yukpa}
\yukpa has lost most of its \setone constructions, mostly employing innovative ones \parencite{meira2006syntactic}.
It preserves the \setone prefixes in the immediate past, where the \gl{1}\gl{s_p_} marker appears to have been extended to \gl{s_a_} verbs, as evidenced by the paradigms in \exref{yukpaintr}.
The loss of split-\gl{s} in favor of the \gl{1}\gl{s_p_} marker is accompanied by the \gl{2}\gl{s_a_} marker \obj{m(ɨ)-} displacing \gl{2}\gl{s_p_} \rc{ə(j)-}, and \gl{1+2} being lost as an inflectional value.
%It has lost the split-\gl{s} system in general

\ex<yukpaintr> \yukpa \parencites[72, 76]{largo2011yukpa}[139]{meira2006syntactic}\\
\begin{tabular}[t]{@{}llll@{}}
& \obj{otum} \qu{to wash self} & \obj{nɨ} \qu{to sleep} & \obj{ata} \qu{to fall}\\
\gl{1} & \obj{j-otum-} & \obj{jɨ-nɨ-} & \obj{j-ata-}\\
\gl{2} & \obj{m-otum-} & \obj{mɨ-nɨ-} & \obj{m-ata-}\\
\gl{3} & \obj{n-otum-} & \obj{nɨ-nɨ-} & \obj{n-ata-}\\
\end{tabular}
\xe
%
The normal intransitive first person marker is \obj{j(ɨ)-}, including those with clear reflexes of \detrz, like \obj{otum} \qu{to wash self} in \exref{yukpaintr}.
The analysis that this is a reflex of the \gl{1}\gl{s_p_} marker \rc{u(j)-} is supported by the same form occurring in \gl{3}>\gl{1} scenarios \exref{yukpatr.yuk-5}.

\pex<yukpatr>\yukpa \parencite[][139]{meira2006syntactic}
\a<yuk-5>
\begingl
\gla aw j-esare//
\glb \gl{1}\gl{pro} \gl{3}>\gl{1}-see//
\glft \qu{S/he saw me.}//
\endgl
\a<yuk-1>
\begingl
\gla aw {\normalfont ∅}-esare//
\glb \gl{1}\gl{pro} \gl{1}>\gl{3}-see//
\glft \qu{I saw it.}//
\endgl
\xe
On the other hand, \gl{1}>\gl{3} scenarios are zero-marked \exref{yukpatr.yuk-1}.
Since \PC \gl{1}\gl{s_a_} \rc{w(ɨ)-} was extended to \gl{1}>\gl{3} scenarios in most languages \parencite[81--82]{gildea1998}, and since it is prone to phonological erosion across the family \pcref{sec:pekodian,sec:waiwaian,sec:akuriyo}, we argue that the zero in \gl{1}>\gl{3} scenarios is in fact the \yukpa reflex of \rc{w-}.
The same zero marking is attested for a single intransitive verb: \obj{to} \qu{to go} \exref{yuk-7}.
Based on regular C-initial verbs taking \obj{jɨ-}, like \qu{to sleep} in \exref{yukpaintr}, it is clear that also C-initial \obj{to} \qu{to go} preserved the old \gl{1}\gl{s_a_} marker \rc{w(ɨ)-}, which was lost due to phonological erosion.

\ex<yuk-7>\yukpa \parencite[][139]{meira2006syntactic}\\
\begingl
\gla aw {\normalfont ∅}-to//
\glb \gl{1}\gl{pro} \gl{1}\gl{s_a_}-go//
\glft \qu{I went.}//
\endgl
\xe

>>>>

All available descriptions of the third Pekodian language, \ikpeng, list \obj{k-} as the only \gl{1}\gl{s_a_} marker \parencites[55]{ikpengpacheco1997}[105]{campetela1997analise}[64]{ikpengpacheco2001}[205]{alves2013verbo}.
However, most \ikpeng cognates of the \arara verbs with \gl{1}\gl{s_a_} \obj{w-} actually do not take \obj{k-}, but rather \obj{ɨ-} or Ø, as shown in \exref{ikpw}.
The exception is \qu{to go}, which has \obj{k-} \exref{ikp-170}.
There is a formally identical \ikpeng cognate of \arara \obj{iptoŋ} \qu{to go down}, but no first person forms are attested \perscommpar{Angela Chagas}.
While there are reflexes of \rc{a[p]} \qu{to be} in \ikpeng, only \rc{eti} forms are attested with first person inflectional prefixes \parencite[401]{gildea2018reconstructing}.

@@

%\ex<ikp-182>\ikpeng \parencite[][320]{chagas2019ikpeng}\\
%\begingl
%\gla ∅-iptoŋ-lan//
%\glb \gl{3}\gl{s_a_}-sit.down-\gl{pst}//
%\glft \qu{She sat down.}//
%\endgl
%\xe
%
While clearly not a regular sound change, \PXin \rc{w} is attested as sometimes being lost in \ikpeng, as evidenced by the correspondences in \cref{tab:pxinw}.
Thus, we suggest that the \ikpeng \gl{1}\gl{s_a_} marker \obj{ɨ-}/Ø is cognate with \arara \gl{1}\gl{s_a_} \obj{w(ɨ)-}, and was subject to phonological erosion of \rc{w}, a phenomenon which will also be seen in other languages discussed in following sections.

%\ex<ikpwloss> %\arara-\ikpeng \obj{w}:Ø correspondences\\
%\begin{tabular}[t]{@{}lll@{}}
%\PXin & \arara & \ikpeng\\
%\rc{wɨna} \qu{\gl{dat}} & \obj{wɨna} & \obj{ɨna}\\
%\rc{wokori} \qu{dog, jaguar} & \obj{wokori} & \obj{akari}\\
%\rc{tawe} \qu{capuchin monkey} & \obj{tawe} & \obj{tae}\\
%\rc{watke} \qu{to defecate} & \obj{watke} & \obj{atke}\\
%\rc{wɨnkɨ} \qu{to sleep} & \obj{wɨnkɨ} & \obj{ɨnkɨ}\\
%& (\obj{o-wɨnkɨ} \qu{sleep!}) & (\obj{j-ɨnkɨ-lɨ} \qu{s/he slept})\\
%\end{tabular}\\
%\parencites[80]{alves2017arara}[164, 67, 150]{ikpengpacheco2001}[9]{desouza2010arara}[68]{souza1993arara}[91]{alves2013verbo}[40]{campetela1997analise} 
%\xe

\input{floats/pxinw}

\input{floats/ppekverbs}

\Cref{tab:ppekverbs} gives an overview of the first person forms of the seven verbs under discussion, along with our \PPek reconstruction.
The form for \rc{ge} \qu{to say} is straightforward to reconstruct; the \bakairi change of \rc{wɨ-} to \obj{u-} is not unexpected given correspondences like \bakairi \obj{udo} \parencite{meira2005southern} from \PC \rc{wɨtoto} \qu{person} \parencite[4]{gildea2007greenberg}.
\qu{to bathe} is straightforwardly reconstructible as \rc{ipɨ}.
For \qu{to be}, \ikpeng \obj{e} is very likely the original vowel, given the \PC form \rc{eti} \pcref{sec:be}.

@@

\pex<hixɨ>
\a<wai-108> \waiwai \parencite[][71]{waiwaihawkins1998}\\
\begingl
\glpreamble wɨɨkekɲe//
\gla wɨ-ka-jakɲe//
\glb \gl{1}-say-\gl{pst}//
\glft \qu{I said.}//
\endgl
\a<hix-118> \hixka \parencite[][124]{hixkaryanaderby1985}\\
\begingl
\glpreamble roxehra nay hamɨ Kaywerye ɨkekonɨ//
\gla ro-ʃe-hɨra n-a-je hamɨ kajwerʲe ɨ-ka-jakonɨ//
\glb \gl{1}-\gl{des}-\gl{neg} \gl{3}-be-\gl{npst}.\gl{uncert} \gl{evid} K. \gl{1}\gl{s_a_}-say-\gl{rem}.\gl{cont}//
\glft \qu{I said (to myself), “Kaywerye evidently doesn't like me”.}//
\endgl
\a<hix-17> \hixka \parencite[][191]{hixkaryanaderby1985}\\
\begingl
\gla ɨ-koroka-no//
\glb \gl{1}>\gl{3}-wash-\gl{imm}//
\glft \qu{I washed him.}//
\endgl
\a<wai-159> \waiwai \parencite[][192]{waiwaihawkins1998}\\
\begingl
\glpreamble wîyesî//
\gla wɨ-jo-jasɨ//
\glb \gl{1}>\gl{3}-boil-\gl{npst}//
\glft \qu{I will boil it.}//
\endgl
\xe
%
Thus, similar to the the case of \ikpeng, \hixka has lost the \rc{w} of the \envr{}{C} allomorph \rc{wɨ-}.
Notably, \textcite{hixkaryanaderby1985} does not see this \obj{ɨ-} as an irregular \gl{1}\gl{s_a_}, but as the regular \gl{1}>\gl{3} marker, because he considers \hixka \obj{ka} \qu{to say} to be transitive, an issue which will be discussed in \cref{sec:say}.

There are three more verbs which did not take innovative \rc{k-} in \PWai; they are shown along with \rc{ka} \qu{to say} in \cref{tab:pwaiverbs}.
While the two roots for \qu{to be} are straightforwardly reconstructible, \qu{to go} is somewhat of a special case.
While \hixka has the expected \obj{ɨ-}, \waiwai seems to have combined innovative \gl{1}\gl{s_a_} \obj{k-} with the old \rc{w-}, an analysis also considered by \textcite[90]{gildea1998}.
Alternatively, this form was influenced by deverbalized forms of \qu{to go}, where a reflex of the \gl{s_a_} class marker \rc{w-} has become fossilized \exref{waiwto}.

@@

\input{floats/pwaiverbs}

%\pex<parpar>
%\a<parcop> Parukotoan \qu{to be}\\
%\begin{tabular}[t]{@{}llll@{}}
%& \PPar & \hixka & \waiwai \\
%\gl{1}\gl{s_a_} & \rc{\emp{w-}eʃi-}/\obj{\emp{w-}ah-} & \obj{\emp{w-}eʃe-}/\obj{\emp{w-}ah-} & \obj{\emp{w-}eeʃi-}/\obj{\emp{w-}a-}\\
%\gl{2}\gl{s_a_} & \rc{m-eʃi-}/\obj{m-ah-} & \obj{m-eʃe-}/\obj{m-ah-} & \obj{m-eeʃi-}/\obj{m-a-}\\
%\gl{1+2}\gl{s_a_} & \rc{t-eʃi-}/\obj{t-ah-} & \obj{t-eʃe-}/\obj{t-ah-} & \obj{t͡ʃ-eeʃi-}/\obj{t-a-}\\
%\gl{3}\gl{s_a_} & \rc{n-eʃi-}/\obj{n-ah-} & \obj{n-eʃe-}/\obj{n-ah-} & \obj{n-eeʃi-}/\obj{n-a-}\\
%\end{tabular}\\
%(\cite[197--198]{hixkaryanaderby1985} and Spike Gildea, p.c.)
%\a<pargo> Parukotoan \qu{to go}\\
%\begin{tabular}[t]{@{}llll@{}}
%& \PPar & \hixka & \waiwai\\
%\gl{1}\gl{s_a_} & \rc{\emp{wɨ-}to-} & \obj{\emp{ɨ-}to-} & \obj{kɨ\emp{w-}to-}\\
%\gl{2}\gl{s_a_} & \rc{mɨ-to-} & \obj{mɨ-to-} & \obj{mɨɨ-to-}\\
%\gl{1+2}\gl{s_a_} & \rc{tɨt-to-} & \obj{tɨ-to-} & \obj{tɨ-to-}/\obj{tɨh-t͡ʃe-}\\
%\gl{3}\gl{s_a_} & \rc{nɨ-to-} & \obj{n-to-} & \obj{nɨɨ-to-}\\
%\end{tabular}\\
%(\cites[69--70, 211]{hixkaryanaderby1985}[179]{waiwaihawkins1998} and Spike Gildea, p.c.)
%\xe
%

@@

As for verbs unaffected by the spread of \rc{t-}, \textcite{meira1998proto} lists the first four verbs in \cref{tab:ptirverbs} as being reconstructible to \PTar with \rc{w-}, for which we offer our reconstructed \PTir forms.

@@

%\pex<ptirw> \PTir \gl{1}\gl{s_a_} \rc{w-} \parencites[112--115]{meira1998proto}[294]{triomeira1999}\\ %[85]{gildea1994akuriyo}
%\begin{tabular}[t]{@{}llll@{}}
%& \PTir & \trio & \akuriyo\\
%\qu{go} & \rc{wɨ-tə(mɨ)-} & \obj{wɨ-tə(mɨ)-} & \obj{ə-təmɨ-} \\% (\obj{wɨ-təemɨ-})\\
%\qu{say} & \rc{wɨ-ka-} & \obj{wɨ-ka-} & \obj{wɨ-ka-} \\
%\qu{come} & \rc{w-əʔepɨ-} & \obj{w-əepɨ-} & \obj{Ø-eepɨ-} \\
%\qu{be} & \rc{w-ae-} & \obj{w-ae-} & Ø\obj{-aʔe-} \\
%& \rc{w-eʔi-} & \obj{w-ei-} & ?-\obj{eʔi-}\\
%\end{tabular}
%\a<aku12> \akuriyo \obj{kəʔ-eepɨ} \qu{we came} \parencite[114]{meira1998proto}
%\a<aku13> \akuriyo \obj{(w)i-toka} \qu{I hit him/her} \parencite[86]{gildea1994akuriyo}
%\a<tricome> \trio \obj{n-epɨ} \qu{s/he came} \parencite[114]{meira1998proto}
%\xe

\input{floats/ptirverbs}

@@

%\begin{table}
%	\centering
%	\caption{\akuriyo \gl{1}\gl{s_a_} markers in Gildea's fieldnotes}
%	\label{tab:aku1sa}
%	\begin{tabular}{@{}lll@{}}
%	\mytoprule
%first person \obj{k-} & first person \obj{t͡ʃ-} \\
%\mymidrule
%\obj{ənɨkɨ} \qu{to sleep} & \obj{eepɨ} \qu{to bathe} \\
%\obj{əməmɨ} \qu{to enter} & \obj{ewai} \qu{to sit down} \\
%\obj{əturu} \qu{to talk} & \obj{etonema} \qu{to lie down} \\
%\obj{əət͡ʃena} \qu{to cry} & \obj{ekɨɨrɨka} \qu{to stay back} \\
%\obj{ətajiŋka} \qu{to run} & \obj{entapo} \qu{to yawn} \\
%\obj{əiwa} \qu{to tremble} &  \\
%\obj{əempa} \qu{to learn} \\
%	\mybottomrule
%	\end{tabular}
%\end{table}

\input{floats/aku1sa.tex}
